Multivariate Skew-Normal Mixture Model for Infant Development Clustering

### 2019-05-13 poster/poster.tex
poster almost done

\usepackage[scale=1.24]{beamerposter} % Use the beamerposter package for laying out the poster

@@

\title{Imputation Methods in Water Quality Modeling} % Poster title

\author{Carter Allen, Dr. Edsel Pena, Dr. John Grego} % Author(s)

\institute{Department of Statistics, University of South Carolina} % Institution(s)

@@

\begin{alertblock}{Objectives}

This project addresses the need for predictive tools to forecast expected concentrations of fecal bacteria Charleston, SC. \textit{Enterococcus faecalis} will be modeled in terms of precipitation in the past 72 hours, location, and tidal stage. Three distict missing/censored data imputation methods are developed and compared to acheive the desired model.

@@

The predictive model is built from data obtained through Charleston Waterkeeper's Recreational Water Quality Monitoring Program (RWQMP), a sampling program that is consistent with protocol stipulated by the South Carolina Department of Health and Environmental Control, as to preserve the compatibility of data collected by the two organizations [3]. The RWQMP measures the concentration of \textit{Enterococcus faecalis} bacteria in units of most probable number of individuals per 100 mL of sample (MPN/100 mL) [3]. Charleston Waterkeeper's records contain both missing and left-cencored (<10) observations. Of the 969 total observations from 2013-2015, 140 are censored and 28 are missing.

@@

\begin{figure}
\includegraphics[width=0.9\linewidth]{SiteLocations.png}
\caption{Locations of Charleston Waterkeeper's Sample Sites}
\end{figure}

@@

%----------------------------------------------------------------------------------------
%	MATERIALS
%----------------------------------------------------------------------------------------

\begin{block}{Exploratory Analysis}
\begin{figure}
\includegraphics[width=0.9\linewidth]{Ln(Ent)vSitevTide.png}
\caption{ln(Ent.) conditioned on tidal stage and sample site}
\end{figure}
\end{block}

@@

\begin{block}{Comparison Criteria}

Lorem ipsum dolor \textbf{sit amet}, consectetur adipiscing elit. Sed laoreet accumsan mattis. Integer sapien tellus, auctor ac blandit eget, sollicitudin vitae lorem. Praesent dictum tempor pulvinar. Suspendisse potenti. Sed tincidunt varius ipsum, et porta nulla suscipit et. Etiam congue bibendum felis, ac dictum augue cursus a. \textbf{Donec} magna eros, iaculis sit amet placerat quis, laoreet id est. In ut orci purus, interdum ornare nibh. Pellentesque pulvinar, nibh ac malesuada accumsan, urna nunc convallis tortor, ac vehicula nulla tellus eget nulla. Nullam lectus tortor, \textit{consequat tempor hendrerit} quis, vestibulum in diam. Maecenas sed diam augue.

@@

\textit{Enterococcus} counts are significantly influenced by precipication, tidal stage, and location. The best performing imputation method of the three tested employs sampling from a truncated normal distribution.

@@

\begin{block}{Imputation Methods}

\begin{enumerate}
\item \textit{A naive approach}: Censored values are sampled from a \( Uniform (0,10) \) and missing values are replaced with the overall mean \textit{Enterococcus} for 2013-2015.
\item \textit{Sampling from a Truncated Normal}: Censored values are sampled from a \(truncNorm()\) distribution, and missing values are predicted by the model made from all non-missing values.
\item \textit{Expected value of a Truncated Normal}: Let \( Y = \) \textit{Enterococcus} concentrations. If \( Y \sim logNormal(\mu,\sigma^2) \implies ln(Y) \sim Normal(\mu,\sigma^2) \), it can be shown that \( E[ln(Y)|a<ln(Y)<b] = \mu+\sigma{\frac{\phi(\frac{a-\mu}{\sigma})-\phi(\frac{b-\mu}{\sigma})}{\Phi(\frac{b-\mu}{\sigma})-\Phi(\frac{a-\mu}{\sigma})}} \). When $a = -\infty$ and $b = ln(10)$,

\end{enumerate}

@@

\begin{block}{Results}

\begin{figure}
\includegraphics[width=0.8\linewidth]{placeholder.jpg}
\caption{Figure caption}
\end{figure}

Nunc tempus venenatis facilisis. Curabitur suscipit consequat eros non porttitor. Sed a massa dolor, id ornare enim:

\begin{table}
\vspace{2ex}
\begin{tabular}{l l l}
\toprule
\textbf{Treatments} & \textbf{Response 1} & \textbf{Response 2}\\
\midrule
Treatment 1 & 0.0003262 & 0.562 \\
Treatment 2 & 0.0015681 & 0.910 \\
Treatment 3 & 0.0009271 & 0.296 \\
\bottomrule
\end{tabular}
\caption{Table caption}
\end{table}

@@

\begin{block}{Conclusion}

Nunc tempus venenatis facilisis. \textbf{Curabitur suscipit} consequat eros non porttitor. Sed a massa dolor, id ornare enim. Fusce quis massa dictum tortor \textbf{tincidunt mattis}. Donec quam est, lobortis quis pretium at, laoreet scelerisque lacus. Nam quis odio enim, in molestie libero. Vivamus cursus mi at \textit{nulla elementum sollicitudin}.

%----------------------------------------------------------------------------------------
%	ADDITIONAL INFORMATION
%----------------------------------------------------------------------------------------

\begin{block}{Additional Information}

Maecenas ultricies feugiat velit non mattis. Fusce tempus arcu id ligula varius dictum. 
\begin{itemize}
\item Curabitur pellentesque dignissim
\item Eu facilisis est tempus quis
\item Duis porta consequat lorem
\end{itemize}

@@

\begin{block}{References}

\nocite{*} % Insert publications even if they are not cited in the poster
\small{\bibliographystyle{unsrt}
\bibliography{sample}\vspace{0.75in}}

@@

\setbeamercolor{block title}{fg=red,bg=white} % Change the block title color

\begin{block}{Acknowledgements}

\small{\rmfamily{Nam mollis tristique neque eu luctus. Suspendisse rutrum congue nisi sed convallis. Aenean id neque dolor. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.}} \\

@@

\begin{alertblock}{Contact Information}

\begin{itemize}
\item Web: \href{http://www.university.edu/smithlab}{http://www.university.edu/smithlab}
\item Email: \href{mailto:john@smith.com}{john@smith.com}
\item Phone: +1 (000) 111 1111
\end{itemize}

@@

\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=0.4\linewidth]{logo.png} & \hfill & \includegraphics[width=0.4\linewidth]{logo.png}
\end{tabular}
\end{center}

>>>>

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage[scale=1.24]{beamerposter} % Use the beamerposter package for laying out the poster

@@

\title{Multivariate Skew-Normal Mixture Model for Infant Development Clustering} % Poster title

\author{Carter Allen$^1$; Brian Neelon, PhD$^1$; Sara E. Benjamin-Neelon, PhD, JD, MPH$^2$} % Author(s)

\institute{$^1$Department of Public Health Sciences, Medical University of South Carolina; $^2$Bloomberg School of Public Health, Johns Hopkins University}

@@

\begin{alertblock}{Abstract}

We propose a novel Bayesian model for infant development patterns that addresses primary research questions in this area while allowing for skewness and correlation of outcomes. Our model is based on finite mixtures of multivariate skew normal (MSN) distributions, where covariates are allowed on both the multivariate outcomes and probability of latent class membership. We also allow for missing outcome data by imputing missing outcomes from their conditional multivariate normal distributions. We demonstrate our method using data from the Nurture study.

@@

A primary goal in infant development research is to identify \textbf{latent development classes} and explain class membership in relation to covariates of interest. It is often also of interest to relate covariates to mean longitudinal growth patterns. Infant development data are inherently correlated longitudinally, often skewed, and frequently missing due to longitudinal attrition and standard practices are ill-suited to addressing these research questions due to their ignorance of one or more of these features of development data.

@@

\begin{block}{Motivation}
\begin{figure}
\includegraphics[width=0.9\linewidth]{bayley_dens.jpg}
\caption{Density of Bayley composite scores for Nurture infants at 3, 6, 9 and 12 months of age.}
\end{figure}
\end{block}

@@

%----------------------------------------------------------------------------------------
%	
%----------------------------------------------------------------------------------------

\begin{block}{Clustering}

A primary concern of our model is with identification of latent infant development clusters. We accomplish this via multinomial logit regression model on cluster membership, which utilizes P\'{o}lya-Gamma data-aumentation to allow for updating of all parameters using Gibbs sampling.

@@

\begin{block}{Conditional Imputation}

We allow for missingness of outcomes in the MSN mixture model by imputing missing values from their conditional multivariate normal distributions. We note that

$$Y_i|X_i,t_i,\boldsymbol\beta,\psi \sim N_k(X_i \boldsymbol\beta + t_i \psi, \boldsymbol\Sigma)$$

This allows us to appeal to standard conditional forms of the multivariate normal distribution. Let $Y_i = [Y^{miss}_{i_{q \times 1}} | Y^{obs}_{i_{k - q \times 1}}]^T$. We have

$$Y_i^{miss}|Y_i^{obs},X_i,t_i,\boldsymbol\beta,\psi \sim N(\mu^{miss},\boldsymbol\Sigma^{miss})$$

where $\mu^{miss}$ and $\Sigma^{miss}$ take standard forms. Each missing outcome is imputed "online", i.e. once per MCMC iteration. This provides more opportunities to explore the parameter space than multiple imputation and avoids multiplicative run-time scaling in $m$, the number of imputations.

@@

We developed a novel Bayesian MSN mixture model, and showed superior performance compared to standard approaches. We applied the MSN mixture model to data from the Nurture study and discovered two distinct development classes characterized by differences in development trajectories and demographics.

@@

\begin{block}{MSN Regression}

We model the effect of covariates on longitudinal development outcomes through the use of a MSN regression model. The MSN distribution can be represented as the superposition of a MN random variable with a latent truncated normal random effect. Let $\mathbf{Y}_{n \times k}$ be the observation matrix such that $Y_{ij}$ is the observation for subject $i$ at timepoint $j$.

$$\mathbf{Y}_{n \times k} = \mathbf{X}_{n \times p}\boldsymbol\beta_{p \times k} + t_{n \times 1}\psi_{1 \times k} + \boldsymbol\epsilon_{n \times k}$$

where $X_i$ is the $1 \times p$ vector of covariate values for subject $i$, $\beta_j$ is the $i \times k$ vector of fixed effects coefficients for timepoint $j$, $t_i \stackrel{iid}{\sim} N_{[0,\infty)}(0,1)$ is a truncated normal random effect, $\psi$ is the vector containing skewness parameters for each timepoint, and $\boldsymbol\epsilon_i \sim N_k(0,\boldsymbol\Sigma_{k \times k})$ is the correlated error term.

@@

\begin{block}{Simulation Study}

\begin{table}[t]

\caption{\label{tab:unnamed-chunk-4}Parameter estimates for n = 1000, k = 4, p = 2, h = 3, v = 3 simulation.}
\centering
\fontsize{20}{20}\selectfont
\begin{tabular}{llllllll}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{2}{c}{Class 1} & \multicolumn{2}{c}{Class 2} & \multicolumn{2}{c}{Class 3} \\
\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8}
Component & Parm. & True & Est. (95\% CrI) & True & Est. (95\% CrI) & True & Est. (95\% CrI)\\
\midrule
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{ }}\\
\hspace{1em}MSN & $\beta_{11}$ & -0.54 & -0.51 (-0.61, -0.41) & -0.19 & -0.22 (-0.35, -0.09) & 1.84 & 1.81 (1.65, 1.99)\\
\hspace{1em}Reg. & $\beta_{12}$ & -1.06 & -1.07 (-1.15, -0.98) & -0.16 & -0.13 (-0.25, -0.02) & 2.03 & 2.09 (1.93, 2.23)\\
\hspace{1em} & $\beta_{13}$ & -1.28 & -1.22 (-1.32, -1.11) & 0.59 & 0.58 (0.46, 0.69) & 1.8 & 1.74 (1.58, 1.9)\\
\hspace{1em} & $\beta_{14}$ & -1.91 & -1.83 (-1.92, -1.74) & -0.2 & -0.12 (-0.25, -0.01) & 1.43 & 1.37 (1.22, 1.52)\\

\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{ }}\\
\hspace{1em} & $\sigma^2_{11}$ & 1 & 0.9 (0.78, 1.03) & 1 & 1.04 (0.88, 1.26) & 1 & 1.39 (1.08, 1.76)\\
\hspace{1em} & $\sigma^2_{12}$ & -0.1 & -0.12 (-0.22, -0.02) & -0.32 & -0.31 (-0.44, -0.18) & 0.08 & 0.27 (0.04, 0.52)\\
\hspace{1em} & $\sigma^2_{13}$ & -0.18 & -0.17 (-0.28, -0.08) & -0.52 & -0.5 (-0.66, -0.38) & 0.78 & 1.19 (0.87, 1.54)\\
\hspace{1em} & $\sigma^2_{14}$ & 0.39 & 0.31 (0.22, 0.43) & 0.04 & 0.05 (-0.07, 0.22) & 0.41 & 0.72 (0.47, 0.99)\\
\hspace{1em} & $\sigma^2_{22}$ & 1 & 0.97 (0.84, 1.12) & 1 & 0.94 (0.77, 1.14) & 1 & 1.18 (0.91, 1.5)\\

\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{ }}\\
\hspace{1em} & $\psi_{1}$ & -0.33 & -0.28 (-0.4, -0.18) & 0.67 & 0.64 (0.5, 0.78) & -1 & -0.91 (-1.12, -0.68)\\
\hspace{1em} & $\psi_{2}$ & -0.33 & -0.21 (-0.34, -0.09) & 0.67 & 0.7 (0.55, 0.84) & -1 & -0.96 (-1.16, -0.76)\\
\hspace{1em} & $\psi_{3}$ & -0.33 & -0.27 (-0.39, -0.14) & 0.67 & 0.72 (0.58, 0.86) & -1 & -0.91 (-1.14, -0.69)\\
\hspace{1em} & $\psi_{4}$ & -0.33 & -0.18 (-0.3, -0.07) & 0.67 & 0.73 (0.6, 0.87) & -1 & -0.93 (-1.13, -0.73)\\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{ }}\\
\hspace{1em}Multi. & $\delta_{11}$ & -0.53 & -0.5 (-0.74, -0.24) & -0.53 & -0.5 (-0.74, -0.24) & -0.53 & -0.5 (-0.74, -0.24)\\
\hspace{1em}Logit & $\delta_{12}$ & 0.44 & 0.28 (0.02, 0.56) & 0.44 & 0.28 (0.02, 0.56) & 0.44 & 0.28 (0.02, 0.56)\\
\hspace{1em} & $\delta_{21}$ & 0.34 & 0.38 (0.09, 0.63) & 0.34 & 0.38 (0.09, 0.63) & 0.34 & 0.38 (0.09, 0.63)\\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{ }}\\
\hspace{1em}Clust. & $\pi_l$ & 0.41 & 0.41 (0.4, 0.42) & 0.36 & 0.36 (0.35, 0.37) & 0.23 & 0.23 (0.22, 0.24)\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
\includegraphics[height=1\linewidth,width=1\linewidth]{beta_trace.pdf}
\caption{Trace plots of regression parameter estimates in simulation}
\end{figure}

\begin{block}{Application Results}

\begin{figure}
\includegraphics[height=0.8\linewidth,width=1\linewidth]{clust_trends.pdf}
\caption{Mean development patterns in each estimated class}
\end{figure}

@@

@@

@@

\begin{alertblock}{Further Resources}

\centering
https://carter-allen.github.io/MVSN-FMM

@@
